Matroids \& testing in families of dependent hypotheses

### 2013-08-20 presentation/presentation.tex
further clean up operations, mostly animations left to do

\begin{block}{$H_1:\mu_1=\mu_2,\; H_2:\mu_2=\mu_3,\;H_3:\mu_1=\mu_3$}
  \begin{enumerate}
  \item Test $H_ {\{1,2,3\}} = H_1\cap H_2 \cap H_3 :
    \mu_1=\mu_2=\mu_3$ at level $\alpha$,
  \item test each $H_1,H_2,H_3$ at level $\alpha$,
  \end{enumerate}
  protects the FWER at level $\alpha$.
\end{block}
\pause
  \begin{itemize}
    \item  essentially closed test, since %; no need to test pairwise
                                %intersection hypotheses since 
        $H_{{i,j}} = H_{\{1,2,3\}}, i\neq j,$

@@

\begin{itemize}
\item $H_i$ depends on $H_I$: $H_{I \cup i} = H_{I}$
\item induces a dependence relation $i \sim_H I$ on index sets $I \in \mathcal{P}(S)$
\item $I$ independent iff for all $i
  \in I$ $i \nsim_H I\setminus i$ (\ie $H_{I} \subsetneq H_{I\setminus i}$)
\pause
\item For pairwise comparisons $I$ independent iff there are no cycles in $\mathscr{G}_I$
\item For general contrast tests $I$ independent iff $\bs{C_I}$ has full
 column rank
\end{itemize} % ends low level
\end{frame}

@@

\item Matroids generalize the structures of graphs and vector spaces,
 \eg cycles, bases, independent sets
\item A matroid $M$ on $S$ endows $\mathcal{P}(S)$ with an algebraic
 structure
\item {\em Cryptomorphism} \ie many ways to define a matroid:
 \begin{itemize}
 \item From a graph perspective: Circuit axioms.
 \item From a vector space perspective: Independence axioms.
 \end{itemize}
%\item There is a dual matroid $M^*$ which can be helpful for proofs and algorithms
\end{itemize} % ends low level
\end{frame}

@@

 \begin{minipage}{.5\textwidth}
   \only<2->{\input{cycle}}
 \end{minipage} 
\end{tabular}

@@

 \begin{minipage}{.5\textwidth}
 \only<1>{\input{independent}}
 \only<2>{\input{cycle}}
 \invisible<1>{\alert{\footnotesize minimally dependent set!}}
 \end{minipage} 
\end{tabular}

@@

\begin{frame}
  \frametitle{}
  \begin{itemize}
  \item Circuits are minimally dependent sets (remove any element to
    get an independent set)
  \item Independent sets are sets for which no subset of elements is a
    circuit
  \item Maximally independent sets are bases (Base axioms)
  \item Cardinality of largest independent subset is the rank of a set
    (Rank axioms)
  \item Maximally dependent sets are called flats (Closure axioms)
  \end{itemize}
\end{frame}

\begin{frame}
\frametitle{Dependence axioms}

\begin{tabular}{ll}
 \begin{minipage}{.4\textwidth}
   \input{dependent}
 \end{minipage} &
 \begin{minipage}{.6\textwidth}
 Relation $\sim_M$ on elements and subsets of $S$,
 if for any $x \in S$, and sets $Y,Z \subseteq S$ holds
 \begin{enumerate}
 \item $y \sim Y$ for all $y \in Y$
 \item if $x \sim_M Y$ and $x \nsim_M Y\setminus y$
   then $y \sim x \cup (Y \setminus y)$
 \item if $x \sim_M Y$ and for all $y \in Y$, $y \sim_M Z$ then $x \sim_M Z$
 \end{enumerate}
 then $\sim_M$ defines a matroid on $S$.
 \end{minipage}
\end{tabular}

@@

\section{Matroids from algebraically dependent hypotheses}
\begin{frame}
 \frametitle{Dependent hypotheses}
 \begin{block}{Equivalence of dependence relations $\sim_H = \sim_M$}
   \begin{itemize}
   \item<1-> $H_i$ depends on $H_I$ iff $H_{I \cup i} = H_{I}$
   \item<2-> Does this mean $\mathscr{H}$ defines a matroid $M$, such that
     $\sim_H = \sim_M$?
   \end{itemize}
 \end{block}
\end{frame}

@@

 \begin{block}{Two-sided pariwise comparisons}
   For a family of two-sided pairwise mean comparisons $H_i$, $i \in
   S$, an elementary hypotheses $H_j$ depends on intersection
   hypotheses $H_I$ iff the corresponding edge $e_j$ is dependent on
   the edge set $\{e_i:i \in I\}$ of the cycle matroid defined by
   $\mathscr{G}_I$.
 \end{block}

 \begin{block}{Two-sided linear contrast tests}
   For a family of two-sided general contrast tests $H_i$, $i \in S$,
   an elementary hypothesis $H_j$ depends on intersection hypotheses
   $H_I$ iff the corresponding contrast vector $\bs{c_j}$ is
   dependent on the column vectors of the matrix $\bs{C}_I$ in the
   vectorial matroid defined by $\bs{C}$ on $\mathbb{R}^m$.
 \end{block}

@@

\begin{frame}
 \frametitle{Matroids from null hypotheses}
 {\Large What is it good for?\\
 \invisible<1>{Structure}}
\end{frame}

\begin{frame}
\frametitle{Maximally independent sets}
 \begin{itemize}
 \item Bases are maximally independent sets of $S$
 \item Rank $\rho I$ is the cardinality of a maximally indepedendent set in $I$
 \item Circuits are minimally dependent sets, \ie each subset of $C$
   with rank $\rho C - 1$ is a maximally independent set of $C$
 \end{itemize}
\end{frame}

\begin{frame}
\frametitle{Maximally dependent sets}
\begin{itemize}
\item The closure of a set $\sigma I = \{i \in S: i \sim I\}$
\item Closed sets {\em flats}: $\mathscr{L} = \{\sigma I: I \subseteq S\}$
\item Hyperplanes are flats of rank $\rho S - 1$
\item All flats are given by closing hyperplanes under intersection
\item Circuits in $M^*$ are the complements of hyperplanes in $M$
\end{itemize} % ends low level
\end{frame}

\begin{frame}{Examples}
\framesubtitle<1>{Independent set}
\framesubtitle<2>{Base}
\framesubtitle<4>{Dependent set}
\framesubtitle<3>{Cycle}
\framesubtitle<5>{Closure}
\only<1>{\input{independent}}
\only<2>{\input{base}}
\only<4>{\input{dependent}}
\only<3>{\input{cycle}}
\only<5>{\input{closure}}
\end{frame}

@@

\item The flats of a matroid are closed under intersection

\item Every elementary hypothesis is a flat
\item For each $I \subseteq S$, $H_I = H_{\sigma I}$
\item Define a stepwise procedure on the flats of $M$

\begin{frame}
 \frametitle{Stepwise test procedure}
 \begin{block}{Multiple testing procedure for algebraically related hypotheses}
   Assume that $\mathscr{H}$ with dependence relation $\sim_H$
   defines a matroid on $M$ on $S$. Then a test procedure that
   rejects elementary hypotheses $H_i$ if all $H_E$, $i \in E$, $E
   \in \mathscr{L}$ are rejected using a local level-$\alpha$ test
   controls the FWER in the strong sense.
 \end{block}

@@

@@

\begin{frame}
\frametitle{Connections to existing procedures}

\textbf{Old wine in new bottles ((c) L. Hothorn)}
\begin{itemize}
\item \cite{shaffer1986modified} represents pairwise comparisons
 by set partitions
\item \cite{bergmann1988improvements} define the concepts of an
 exhaustive sets
\item \cite{bernhard1991computergestuetzte} shows properties of
 exhaustive sets that are basically {\em closure axioms} of a matroid
 and provides algorithms to check each intersection hypotheses if is
 a flat using rank function.
\item \cite{westfall1997multiple} generates flats using dependence
 arguments in vector spaces.
\item \cite{weichert2000robuste} suggest an MTP based on clique
 decomposition of the graph, cliques in the graph are flats
\item \cite{westfall2007multiple} provide a branch and bound algorithm
 to generate all flats. They note that for pairwise comparisons all
 exhaustive hypotheses are {\em partition hypotheses}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Matroid representation}
 \begin{itemize}
 \item proofs are shorter (more elegant),
 \item efficient algorithms exist to find the flats of matroids,
 \item {\em Scum principle}\cite{welsh2010matroid} everything
   important happens at the top (hyperplanes)
 \item we may define procedures using independent sets,
 \item \ldots{}
 \end{itemize}
\end{frame}

@@

 \end{minipage} &
 \begin{minipage}{.6\textwidth}
   \begin{itemize}
   \item Provides weights $w_{i,I}$ for each index $i$ and each
     non-empty subset $I \subset S$
   \item Weights are used to define a closed test procedure of
     weighted intersection hypotheses tests
   \item We have $\sum_{i \in S} w_{i,I} \leq 1$ (conservative),
   \item and $w_{i,I} \leq w_{i,J}$ for $J \subseteq I$ (consonance).
   \end{itemize}
 \end{minipage}
\end{tabular}
\end{frame}

@@

\begin{frame}
 \frametitle{Fixed sequence test of pairwise mean comparisons}
\begin{itemize}
\item {\bf Picture of hierarchical test hereabouts}
\end{itemize}

@@

\begin{enumerate}
\item Pretest using weighted Bonferroni or conventional graphical
 approach $R = \{i \in S: H_i \textrm{ rejected by pretest}\}$
\item Compute all matroid hyperplanes $\mathscr{H}$ for the matroid on
 the remaining hypotheses $I$
\item $R \gets R \cup \{i: p_i \leq \min_J w_{i,J} \alpha, J \in \mathscr{H}\}$
\item $A \gets \{\bigcup J: \min_j (p_j / w_{j,J}) > \alpha\}$
\item $I \gets I \setminus R$
\item Refine: $\mathscr{H} \gets \{J\in \mathscr{L}: \rho J = \rho
 \mathscr{H} - 1\}$
\item Prune: $\mathscr{H} \gets \{J \in \mathscr{H}: J \cap R = \emptyset\}$
\item Repeat until $\rho \mathscr {H} = 0$
\item Reject $R$ and $I \setminus A$
\end{enumerate}
\end{frame}
\section{Matriods from general families of hypotheses}

\begin{frame}
\frametitle{General Results}
\begin{itemize}
\item We give sufficient (and probably necessary) conditions for a
 general family of hypotheses $\mathscr{H}$ to define a matroid
\item We give sufficient (and probably necessary) conditions on
 $\mathscr{H}$ for $\mathscr{L}$ to be the set of strictly exhaustive
 sets
\item Systems with one sided hypotheses, and parallel hypotheses (\eg
 combined non-inferiority/superiority) do not give matroids with
 $\mathscr{L}$ all strictly exhaustive
\end{itemize}

@@

\begin{frame}
 \frametitle{Computation of $\mathscr{F}$}
 The question remains how to compute $\mathcal{F}$. For this observe
that the $h_i$ define a hyperplane configuration and $\mathcal{F}$ is
just the face lattice of the corresponding oriented
matroid. Consequently, $\mathcal{F}$ can be found using algorithms
for arrangement construction from polyhedral programming
\cite{Avis-KF-92,Avis-KF-96,Ferrez-KF-Liebling-01} which are
polynomial in time and input size and compact in space!

>>>>

\begin{block}{$H_1:\mu_1=\mu_2,\; H_2:\mu_2=\mu_3,\;H_3:\mu_1=\mu_3$}
  \begin{enumerate}
  \item Test $H_ {\{1,2,3\}} = H_1\cap H_2 \cap H_3 :
    \mu_1=\mu_2=\mu_3$ at level $\alpha$,
  \item test each $H_1,H_2,H_3$ at level $\alpha$,
  \end{enumerate}
  protects the FWER at level $\alpha$.
\end{block}
\pause
  \begin{itemize}
    \item  essentially closed test, since %; no need to test pairwise
                                %intersection hypotheses since 
        $H_{\{i,j\}} = H_{\{1,2,3\}}, i\neq j,$

@@

\begin{itemize}
\item $H_i$ \alert{depends} on $H_I$: $H_{I \cup i} = H_{I}$
\item induces a \alert{dependence relation} $i \alert{\sim_H} I$ on index sets $I \in \mathcal{P}(S)$
\item $I$ \alert{independent} iff for all $i
  \in I$ $i \nsim_H I\setminus i$ (\ie $H_{I} \subsetneq H_{I\setminus i}$)
\pause
\item For pairwise comparisons $I$ independent iff there are no cycles
  in the corresponding graph
\item For general contrast tests $I$ independent iff $\bs{C_I}$ has full
 column rank
\end{itemize} % ends low level
\end{frame}

@@

\item Matroids generalize the structures of graphs and vector spaces,
 \eg cycles, bases, independent sets
\item A matroid $M$ on $S$ endows $\mathcal{P}(S)$ with an algebraic
 structure
\item \alert{Cryptomorphism} \ie many ways to define a matroid:
 \begin{itemize}
 \item From a graph perspective: Circuit axioms.
 \item From a vector space perspective: Independence axioms.
 \end{itemize}
%\item There is a dual matroid $M^*$ which can be helpful for proofs and algorithms
\end{itemize} % ends low level
\end{frame}

@@

 \begin{minipage}{.5\textwidth}

** TODO shorten animation

   \only<2->{\input{cycle}}
 \end{minipage} 
\end{tabular}

@@

 \begin{minipage}{.5\textwidth}

** TODO animate independent sets

 \input{independent}
% \only<2>{\input{cycle}}
% \invisible<1>{\alert{\footnotesize minimally dependent set!}}
 \end{minipage} 
\end{tabular}

@@

\begin{frame}
  \frametitle{Matroid structure}

** TODO add some pictures detailing general structure

  \begin{itemize}
  \item \alert{Circuits} are minimally dependent sets (remove any element to
    get an independent set)
  \item \alert{Independent sets} are sets for which no subset of elements is a
    circuit
  \item Maximally independent sets are \alert{bases} (Base axioms)
  \item Cardinality of largest independent subset is the \alert{rank} of a set
    (Rank axioms)
  \item Maximally dependent sets are called \alert{flats} (Closure axioms)
  \end{itemize}
\end{frame}

\begin{frame}
\frametitle{Dependence axioms}
** TODO animate dependence
\begin{tabular}{ll}
 \begin{minipage}{.4\textwidth}
   \input{dependent}
 \end{minipage} &
 \begin{minipage}{.6\textwidth}
 Relation $\sim_M$ on elements and subsets of $S$,
 if for any $j \in S$, and sets $I,K \subseteq S$ holds
 \begin{enumerate}
 \item $i \sim_M I$ for all $i \in I$
 \item if $j \sim_M I$ and $j \nsim_M I\setminus \{i\}$
   then $i \sim \{j\} \cup (I \setminus \{i\})$
 \item if $j \sim_M I$ and for all $i \in I$, $i \sim_M K$ then $j \sim_M K$
 \end{enumerate}
 then $\sim_M$ defines a matroid on $S$.
 \end{minipage}
\end{tabular}

@@

\section{Matroids from algebraically dependent hypotheses}
\begin{frame}
 \frametitle{Dependent hypotheses}
 \begin{block}{Equivalence of dependence relations $\sim_H = \sim_M$}
   \begin{itemize}
   \item<1-> $H_i$ depends on $H_I$ iff $H_{I \cup i} = H_{I}$
   \item<2-> Does this mean a family of algebraically dependent
     hypotheses defines a matroid $M$, such that $\sim_H = \sim_M$?
   \end{itemize}
 \end{block}
\end{frame}

@@

 \begin{block}{Two-sided pariwise comparisons}
   For a family of two-sided pairwise mean comparisons $H_i$, $i \in
   S$, an elementary hypotheses $H_j$ depends on intersection
   hypotheses $H_I$ iff the corresponding edge $e_j$ is dependent on
   the edge set $\{e_i:i \in I\}$ of the \alert{cycle matroid} defined by
   the corresponding graph.
 \end{block}

 \begin{block}{Two-sided linear contrast tests}
   For a family of two-sided general contrast tests $H_i$, $i \in S$,
   an elementary hypothesis $H_j$ depends on intersection hypotheses
   $H_I$ iff the contrast vector $\bs{c_j}$ is a linear combination of
   column vectors of contrast matrix $\bs{C}_I$ which corresponds to
   the \alert{vectorial matroid} defined by $\bs{C}$ on $\mathbb{R}^m$.
 \end{block}

@@

\begin{frame}
 \frametitle{Matroids from algebraically dependent hypotheses}
 {\Large What is it good for?\\
 \invisible<1>{Structure}}
\end{frame}

% \begin{frame}
% \frametitle{Maximally independent sets}
%  \begin{itemize}
%  \item Bases are maximally independent sets of $S$
%  \item Rank $\rho I$ is the cardinality of a maximally indepedendent set in $I$
%  \item Circuits are minimally dependent sets, \ie each subset of $C$
%    with rank $\rho C - 1$ is a maximally independent set of $C$
%  \end{itemize}
% \end{frame}

% \begin{frame}
% \frametitle{Maximally dependent sets}
% \begin{itemize}
% \item The closure of a set $\sigma I = \{i \in S: i \sim I\}$
% \item Closed sets {\em flats}: $\mathscr{L} = \{\sigma I: I \subseteq S\}$
% \item Hyperplanes are flats of rank $\rho S - 1$
% \item All flats are given by closing hyperplanes under intersection
% \item Circuits in $M^*$ are the complements of hyperplanes in $M$
% \end{itemize} % ends low level
% \end{frame}

% \begin{frame}{Examples}
% \framesubtitle<1>{Independent set}
% \framesubtitle<2>{Base}
% \framesubtitle<4>{Dependent set}
% \framesubtitle<3>{Cycle}
% \framesubtitle<5>{Closure}
% \only<1>{\input{independent}}
% \only<2>{\input{base}}
% \only<4>{\input{dependent}}
% \only<3>{\input{cycle}}
% \only<5>{\input{closure}}
% \end{frame}

@@

\item Flats are maximally dependent sets \ie $j \nsim I \forall j \in
  S\setminus I$
\item The flats of a matroid are closed under intersection
\item $S$ (global null hypothesis) is a flat
\item Every single element of $S$ (elementary hypothesis) is a flat
\item For each $I \subseteq S$, the $H_I$ as a subset of the parameter
  space is identical to the intersection hypotheses defined by it
  closure $H_{\sigma I}$ (\ie flats correspond to weakly exhaustive
  sets \cite{bergmann1988improvements})
\item Define a stepwise procedure on the flats of $M$
\end{itemize} % ends low level
\end{frame}

\begin{frame}
 \frametitle{Stepwise test procedure}
 \begin{block}{Multiple testing procedure for algebraically related hypotheses}
   Assume that a family of null hypotheses $\mathscr{H}$ with
   dependence relation $\sim_H$ defines a matroid $M$ on $S$. Then a
   test procedure that rejects elementary hypotheses $H_i$ if for all
   flats $E \subseteq S$ with $i \in E$, $H_E$ is rejected using a
   local level-$\alpha$ test, controls the FWER in the strong sense.
 \end{block}

@@

\begin{frame}
\frametitle{Old wine in new bottles?}

\begin{itemize}
\item \cite{shaffer1986modified} represents pairwise comparisons
 by set partitions
\item \cite{bergmann1988improvements} define the concepts of an
 exhaustive sets
\item \cite{bernhard1991computergestuetzte} shows properties of
 exhaustive sets that are basically \alert{closure axioms} of a matroid
 and provides algorithms to check each intersection hypotheses if is
 a flat using rank function.
\item \cite{westfall1997multiple} generates flats using dependence
 arguments in vector spaces.
\item \cite{weichert2000robuste} suggest an MTP based on clique
 decomposition of the graph, cliques in the graph are flats
\item \cite{westfall2007multiple} provide a branch and bound algorithm
 to generate all flats. They note that for pairwise comparisons all
 exhaustive hypotheses are \alert{partition hypotheses}
\item \cite{brannath2010shortcuts} provide shortcuts for stepwise
  procedures on \eg the set of exhaustive index sets
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Added benefits}
 \begin{itemize}
 \item Many results about exhaustive sets follow directly from
   corresponding results in matroid theory.
 \item Efficient algorithms to find the flats of specific and general
   matroids are available from matroid theory.
 \item \alert{Scum principle} \cite{welsh2010matroid} everything
   important happens at the top \ie it is sufficient to find the
   \alert{hyperplanes} of the corresponding matroid.
 \item The structure of flats provides some intuition to procedures
   that operate on exhaustive sets. 
 \end{itemize}
\end{frame}

@@

@@

 \end{minipage} &
 \begin{minipage}{.6\textwidth}
   \begin{itemize}
   \item Provides weights $w(I)_i$ for each index $i$ and each
     non-empty subset $I \subset S$
   \item Weights are used to define a closed test procedure of
     weighted intersection hypotheses tests
   \item We have $\sum_{i \in S} w(I)_{i} \leq 1$ (conservative),
   \item and $w(I)_{i} \leq w(J)_{i}$ for $J \subseteq I$ (consonance).
   \end{itemize}
 \end{minipage}
\end{tabular}
\end{frame}

@@

\begin{frame}
 \frametitle{Fixed sequence test of pairwise mean comparisons}

** TODO Picture of hierarchical test hereabouts

@@

\begin{enumerate}
\item Pretest using consonant weighted closed test procedure (\eg
  graphical approach), let $R = \{i \in S: H_i \textrm{ rejected by pretest}\}$
\item Compute all matroid hyperplanes $\mathscr{L}$ for the matroid
  restricted to the remaining hypotheses $I$
\item $R \gets R \cup \{i: p_i \leq \min\{ w(J)_{i}\alpha: J \in
  \mathscr{L} \} \}$
\item $A \gets \{\bigcup J: \min\{p_j / w(J)_{j}: j \in J\} > \alpha,
J \in \mathscr{L} \}$
\item $I \gets I \setminus R$
\item Compute hyperplanes of next smaller rank: $\mathscr{L} \gets \{J\in \mathscr{L}: \rho J = \rho
 \mathscr{L} - 1\}$
\item Remove interval of rejected hypotheses: $\mathscr{L} \gets \{J \in \mathscr{L}: J \cap R = \emptyset\}$
\item Reduce to interval of not accepted: $\mathscr{L} \gets \{J \in \mathscr{L}: L \nsubseteq A\}$
\item Repeat until $\rho \mathscr{L} = 0$
\item Reject $R$ and $I \setminus A$
\end{enumerate}
\end{frame}
\section{Matriods from general families of hypotheses}

\begin{frame}
\frametitle{General Results}
\begin{itemize}
\item We give sufficient necessary conditions for a general family of
  hypotheses $\mathscr{H}$ to define a matroid
\item Systems with one sided hypotheses, and parallel hypotheses (\eg
 combined non-inferiority/superiority) do not give matroids with
 $\mathscr{L}$ all strictly exhaustive
\item It is an open question under which conditions the set of
  strictly exhaustive sets is given by the flats of a matroid
\end{itemize}

@@

\begin{frame}
 \frametitle{Computation of $\mathscr{F}$}
 The question remains how to compute $\mathcal{F}$. For this observe
that the $h_i$ define a hyperplane configuration and $\mathcal{F}$ is
just the face lattice of the corresponding oriented
matroid. Consequently, $\mathcal{F}$ can be found using algorithms
for arrangement construction from polyhedral programming
\cite{avis1992pivoting,avis1996reverse,ferrez2001cuts} which are
polynomial in time and input size and compact in space!
